---
title: "Detection Capability: `{sherloq}` R Package"
output: rmdformats::readthedown
vignette: >
  %\VignetteIndexEntry{Detection Capability: `{sherloq}` R Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**`{sherloq}` is an R package for detection capability calculations. It contains 6 functions based on CLSI EP17 guidelines**:

- Limit of Blank (LoB):
    1. Classical (non-parametric & parametric)
- Limit of Detection (LoD)
    2. Classical
    3. Precision profile
    4. Probit
- Limit of Quantitation (LoQ)
    5. Functional sensitivity
    6. Total error

This vignette includes demonstrations of each `{sherloq}` function using CLSI EP17 example data, which can be found in the CLSI EP17 appendix. The LoB, LoD_precision_profile, LoD_probit, and LoQ_total_error functions are also demonstrated using MTM Calibration DEV data obtained from the DEV_Linearity_Precision_LoD.Rmd file in the mtm.calibration package. The exact line numbers from which these data are pulled is included in each example below.

```{r setup}
# load sherloq package
library(sherloq)
```


# LoB

**Example data (CLSI EP17 Appendix A):**

```{r, echo = FALSE}

LoB_df <- data.frame(Reagent = c(rep(1, 12*5), rep(2, 12*5)),
                       Day = rep(rep(c(1, 2, 3), each = 4, times = 10)),
                       Sample = rep(LETTERS[1:5], each = 12, times = 2),
                       Replicate = rep(c(1, 2, 3, 4), times = 3*5*2),
                       Measurement = c(2.6, -.8, 5.5, 6.0,
                                   4.5, .6, -2.3, 3.4,
                                   5.9, 7.6, 4.1, -1.4,
                                   1.0, 2.9, 4.9, 8.0,
                                   6.9, 5.0, 3.4, 1.2,
                                   6.5, 5.6, -2.2, 2.3,
                                   -4.4, -3.4, 7.0, 6.9,
                                   4.3, 3.2, -1.4, 4.2,
                                   5.9, 7.6, 3.8, 5.8,
                                   1.5, -1.9, 5.1, 5.7,
                                   4.1, 4.5, -.6, .5,
                                   5.4, 7.6, 4.4, 6.6,
                                   1.2, -.7, 6.1, 5.1,
                                   4.8, 3.3, -2.8, -1.4,
                                   8.7, 3.6, 5.1, 3.5,
                                   4.6, 4.1, 1.6, 3.7,
                                   2.2, .7, 4.6, 2.6,
                                   1.1, -4.4, .9, .7,
                                   9.2, 8.3, 4.8, 5.4,
                                   4.8, 6.3, 5.4, 9.6,
                                   7.7, 3.1, 6.1, 10.0,
                                   6.1, 3.2, 3.9, 1.4,
                                   3.1, 4.1, 1.0, 3.4,
                                   .1, .4, 2.9, -1.6,
                                   4.0, 11.5, 4.5, 3.6,
                                   4.4, 6.8, 7.1, 4.2,
                                   3.7, 3.7, 5.3, 4.5,
                                   4.0, 6.2, -.2, 2.3,
                                   1.6, 2.6, 6.4, 5.7,
                                   4.2, 3.7, 1.4, 1.5))

head(LoB_df, 10) |> 
  knitr::kable()
```

Use the `{sherloq}` LoB function:

```{r}
results <- LoB(df = LoB_df, 
               col_lot = "Reagent", 
               col_sample = "Sample", 
               col_value = "Measurement"
               )

# get LoB values
results$LoB_values
```

<br>

**MTM Calibration DEV data:**  
(obtained from DEV_Linearity_Precision_LoD.Rmd, line 63)

```{r, echo = FALSE}
data(mtm.cal_LoB_dat)

LoBdat <- LoBdat |>
  subset(select = c(IntendedVAF, ObsVAF, Reagent, Rep, Day, SampleCall, PlasmaSampleNumber))

head(LoBdat, 10) |> 
  knitr::kable()
```

Use the `{sherloq}` LoB function:

```{r}
results <- LoB(df = LoBdat, 
               col_lot = "Reagent", 
               col_sample = "PlasmaSampleNumber", 
               col_value = "ObsVAF")

# get LoB values
results$LoB_values
```

Additional features of this function:

- If there are more than three reagent lots, the function automatically pools them (per CLSI guidelines)
    - Can set `col_lot = NULL` to pool all lots
    - Can set `always_sep_lots = TRUE` to evaluate all lots separately
- Can set `parametric = TRUE` to use the parametric approach
    - Checks for normal distribution (an assumption of the parametric approach)
- Can change alpha level (e.g. `alpha = 0.01`) 



# LoD (Classical)

**Example data (CLSI EP17 Appendix A):**

```{r, echo = FALSE}
LoD_C_df <- data.frame(Reagent = c(rep(1, 12*5), rep(2, 12*5)),
                        Day = rep(rep(c(1, 2, 3), each = 4, times = 10)),
                        Sample = rep(LETTERS[1:5], each = 12, times = 2),
                        Replicate = rep(c(1, 2, 3, 4), times = 3*5*2),
                        Measurement = c(21.0, 22.8, 28.2, 25.9,
                                    26.4, 28.3, 20.7, 21.9,
                                    24.7, 22.5, 28.5, 29.2,
                                    13.3, 12.6, 18.2, 14.7,
                                    17.8, 14.0, 14.1, 12.5,
                                    11.3, 12.2, 16.2, 13.9,
                                    12.8, 12.9, 17.4, 16.0,
                                    15.9, 14.1, 11.3, 9.4,
                                    10.6, 13.6, 17.6, 14.9,
                                    17.3, 19.2, 21.5, 22.2,
                                    24.1, 25.8, 16.0, 16.4,
                                    24.9, 23.8, 22.1, 26.1,
                                    19.2, 22.7, 28.3, 26.2,
                                    25.1, 30.3, 23.4, 19.2,
                                    26.3, 23.1, 27.5, 30.1,
                                    22.0, 22.5, 21.8, 22.1,
                                    20.3, 21.0, 25.3, 26.0,
                                    27.2, 25.1, 25.3, 25.3,
                                    15.6, 21.2, 14.8, 14.9,
                                    16.0, 15.8, 21.6, 22.8,
                                    15.3, 18.7, 18.3, 19.5,
                                    13.0, 15.9, 9.0, 7.0,
                                    13.4, 8.5, 16.3, 18.1,
                                    12.4, 11.1, 11.3, 10.1,
                                    18.8, 17.6, 14.1, 14.9,
                                    19.2, 15.8, 19.8, 21.4,
                                    18.0, 18.0, 19.6, 23.1,
                                    32.9, 30.4, 29.4, 27.6,
                                    27.7, 30.6, 31.4, 30.4,
                                    32.5, 28.9, 29.8, 35.1))
head(LoD_C_df, 10) |> 
  knitr::kable()
```


Use the `{sherloq}` LoD_classical function:

```{r}
results <- LoD_classical(df = LoD_C_df, 
                         col_lot = "Reagent", 
                         col_sample = "Sample", 
                         col_value = "Measurement", 
                         LoB = 7.5)

# get LoD values
results$LoD_values
```

Additional features of this function:

- If there are more than three reagent lots, the function automatically pools them (per CLSI guidelines)
    - Can set `col_lot = NULL` to pool all lots
    - Can set `always_sep_lots = TRUE` to evaluate all lots separately
- Can change beta level (e.g. `beta = 0.01`) 

<br>

# LoD (Precision Profile)

**Example data (CLSI EP17 Appendix B):**

```{r, echo = FALSE}
LoD_PP_df <- data.frame(Reagent = c(rep(1, 6), rep(2, 6)),
                       Sample = rep(LETTERS[1:6], 2),
                       Mean = c(.69, 1.42, 2.65, 4.08, 6.08, 10.36,
                                .78, 1.73, 2.89, 3.82, 6.33, 10.92),
                       SD_within_lab = c(.39, .39, .46, .55, .64, 1.12, 
                                         .29, .54, .55, .63, .82, 1.38))
LoD_PP_df |> 
  knitr::kable()
```

Use the `{sherloq}` LoD_precision_profile function:

```{r}
results <- LoD_precision_profile(df = LoD_PP_df, 
                                 col_lot = "Reagent", 
                                 col_sample = "Sample",
                                 col_avg = "Mean", 
                                 col_sd_wl = "SD_within_lab", 
                                 LoB = 0.51, 
                                 n_measures = 80*6)

# get LoD values
results$LoD_values

# get model coefficients from model objects
results$model |>
  lapply(function(x) summary(x)$coef)

# see head of table to recreate plot
results$plot_data |>
  head(10) |>
  knitr::kable()

```

<br>

**MTM Calibration DEV data:**  
(obtained from DEV_Linearity_Precision_LoD.Rmd, line 390)

```{r, echo = FALSE}
data(mtm.cal_LoD_precision_profile_dat)

LoDdat_PP <- LoDdat_PP |>
  subset(DataType == "Mixed Model" & PrecType == "Reproducibility",
         select = -c(PrecType, DataType)) |>
  `row.names<-`(NULL)

LoDdat_PP |> knitr::kable()
```

Use the `{sherloq}` LoD_precision_profile function:

```{r}
results <- LoD_precision_profile(df = LoDdat_PP, 
                                 col_lot = NULL, 
                                 col_sample = "Intended",
                                 col_avg = "Observed", 
                                 col_sd_wl = "SD",
                                 LoB = 0, 
                                 n_measures = 149)

# get LoD values
results$LoD_values

# get model coefficients from model objects
results$model |>
  lapply(function(x) summary(x)$coef)

# see head of data to recreate plot
results$plot_data |>
  head(10) |>
  knitr::kable()

```

Additional features of this function:

- If there are more than three reagent lots, the function automatically pools them (per CLSI guidelines)
    - Can set `col_lot = NULL` to pool all lots
    - Can set `always_sep_lots = TRUE` to evaluate all lots separately
- Can change beta level (e.g. `beta = 0.01`) 
- Can select a specific model for the precision profile or default to the model with the lowest AIC


<br>

# LoD (Probit)

**Example data (CLSI EP17 Appendix C):**

```{r, echo = FALSE}
LoD_P_df <- data.frame(Concentration = rep(c(0, .006, .014, .025, .05, .15, .3, .5), each = 3),
                       Reagent = rep(c(1, 2, 3), times = 8),
                       Observed_Positives = c(0, 0, 0,
                                              11, 12, 22,
                                              15, 22, 31,
                                              23, 28, 27,
                                              29, 32, 32,
                                              32, 32, 32,
                                              32, 32, 32,
                                              32, 32, 32),
                       Total_Calls = c(22, 22, 22,
                                       30, 30, 34,
                                       30, 30, 34,
                                       rep(32, 15)))

head(LoD_P_df, 10) |> 
  knitr::kable()
```

Use `{sherloq}` LoD_probit function:

```{r}
results <- LoD_probit(df = LoD_P_df, 
                      col_lot = "Reagent", 
                      col_true_conc = "Concentration", 
                      col_obs_pos = "Observed_Positives", 
                      col_tot = "Total_Calls", 
                      LoB = 0)

# get LoD values
results$LoD_values

# get model coefficients from model objects
results$model |>
  lapply(function(x) summary(x)$coef)

# see head of table to recreate plot
results$plot_data |>
  head(10) |>
  knitr::kable()
```


**MTM Calibration DEV data:**  
(obtained from DEV_Linearity_Precision_LoD.Rmd, line 332)

```{r, echo = FALSE}
data(mtm.cal_LoD_probit_dat)

LoDdat_P <- LoDdat_P |>
  subset(select = c(IntendedVAF, ObsVAF, Reagent, SampleCall))

head(LoDdat_P, 10) |> 
  knitr::kable()
```

This data is in a different format, but the LoD_probit function can also handle this data structure. We just specify `col_01` instead of `col_obs_pos` and `col_tot`.

The DEV study code does not separate reagent lots, so we will pool all lots together with col_lot = NULL.

Use the `{sherloq}` LoD_probit function:

```{r}
results <- LoD_probit(df = LoDdat_P, 
                      col_lot = NULL, 
                      col_true_conc = "IntendedVAF",
                      col_01 = "SampleCall", 
                      LoB = 0)

# get LoD values
results$LoD_values

# get model coefficients from model objects
results$model |>
  lapply(function(x) summary(x)$coef)

# see head of table to recreate plot
results$plot_data |>
  head(10) |>
  knitr::kable()

```

According to CLSI guidelines, LoD should be calculated separately for each lot. What happens if we separate by lot?

```{r}
results <- LoD_probit(df = LoDdat_P, 
                      col_lot = "Reagent", 
                      col_true_conc = "IntendedVAF",
                      col_01 = "SampleCall",
                      LoB = 0, 
                      log10_trans = TRUE)

# get LoD values
results$LoD_values

# get model coefficients from model objects
results$model |>
  lapply(function(x) summary(x)$coef)

# see head of table to recreate plot
results$plot_data |>
  head(10) |>
  knitr::kable()
```

Additional features of this function:

- If there are more than three reagent lots, the function automatically pools them (per CLSI guidelines)
    - Can set `col_lot = NULL` to pool all lots
    - Can set `always_sep_lots = TRUE` to evaluate all lots separately
- Can change beta level (e.g. `beta = 0.01`) 
- Can set `log10_trans = TRUE` to perform a log transformation of the concentration values before fitting the probit models
    - CLSI guidelines indicate that this may lead to better model fit


<br>

# LoQ (Functional Sensitivity)

**Example data (CLSI EP17 Appendix D1):**

```{r, echo = FALSE}
LoQ_F_df <- data.frame(Reagent = rep(c(1, 2), each = 9),
                       Sample = rep(1:9, times = 2),
                       Mean = c(.04, .053, .08, .111, .137, .164, .190, .214, .245,
                                .041, .047, .077, .106, .136, .159, .182, .205, .234),
                       SD_within_lab = c(.016, .016, .016, .017, .014, .012, .011, .016, .013,
                              .018, .014, .012, .019, .016, .015, .015, .016, .014))

head(LoQ_F_df, 10) |> knitr::kable()
```

Use the `{sherloq}` LoQ_total_error function:

```{r}
results <- LoQ_functional(df = LoQ_F_df, 
                          col_lot = "Reagent", 
                          col_avg = "Mean", 
                          col_sd_wl = "SD_within_lab", 
                          target_cv = 10)

# get LoQ values
results$LoQ_values

# get model coefficients from model objects
results$model |>
  lapply(function(x) summary(x)$coef)

```

Additional features of this function:

- If there are more than three reagent lots, the function automatically pools them (per CLSI guidelines)
    - Can set `col_lot = NULL` to pool all lots
    - Can set `always_sep_lots = TRUE` to evaluate all lots separately
- Can set `model = "power w intercept"` to fit a power function model with an intercept 
- NLS starting coefficients are determined within the function
    - Can supply starting coefficients with `coeff_start`


<br>

# LoQ (Total Error)

**Example data (CLSI EP17 Appendix D2):**

```{r, echo = FALSE}
LoQ_TE_df <- data.frame(Reagent = c(rep(1, 9*5), rep(2, 9*5)),
                         Day = rep(rep(c(1, 2, 3), each = 3, times = 10)),
                         Sample = rep(c(1, 2, 3, 4, 5), each = 9, times = 2),
                         Reference_Value = rep(c(38.2, 47.1, 44.7, 36.5, 42.8), each = 9, times = 2),
                         Replicate = rep(c(1, 2, 3), times = 3*5*2),
                         Measurement = c(36.7, 37.9, 38.3, 36.8, 33.5, 39.2, 41.3, 37.9, 34.9,
                                 49.9, 50, 48.1, 47.8, 43.9, 45.6, 45.4, 51.5, 45.8,
                                 46.1, 43.1, 39.4, 47.3, 45.8, 44.8, 44.6, 47.3, 38.9,
                                 33.3, 34.2, 34.5, 43.1, 34, 37.1, 35.3, 32.4, 36, 42.9,
                                 41.8, 43.8, 46.3, 43.3, 46, 42.6, 41.4, 42.8, 38.5, 41,
                                 43.2, 36.8, 42.1, 35.8, 36.8, 44.1, 39.5, 45.8, 47.8,
                                 46.6, 46.9, 51.3, 50.5, 44.3, 47.5, 52.4, 46.7, 43.6,
                                 42.4, 46.5, 47.9, 42.7, 42.1, 43.4, 44.7, 35.5, 40, 34,
                                 32.9, 33.1, 38.6, 36.2, 41.4, 33, 42, 44.1, 43.2, 46.6,
                                 45.5, 43.5, 41.4, 48.2, 45.7))

head(LoQ_TE_df, 10) |> knitr::kable()
```

Use the `{sherloq}` LoQ_total_error function:

```{r}
results <- LoQ_total_error(df = LoQ_TE_df, 
                           col_lot = "Reagent", 
                           col_sample = "Sample", 
                           col_ref = "Reference_Value", 
                           col_value = "Measurement", 
                           accuracy_goal = 21.6)

# get LoQ values
results$LoQ_values

# get total error table
results$total_error_table |> 
  knitr::kable()

```

**MTM Calibration DEV data:**  
(obtained from DEV_Linearity_Precision_LoD.Rmd, line 590)

```{r, echo = FALSE}
data(mtm.cal_LoQ_total_error_dat)

LoQdat <- LoQdat |>
  subset(select = c(IntendedVAF, ObsVAF, Reagent))

head(LoQdat, 10) |> knitr::kable()
```

Use the `{sherloq}` LoQ_total_error function:

```{r}
results <- LoQ_total_error(df = LoQdat, 
                           col_lot = NULL, 
                           col_sample = "IntendedVAF",
                           col_ref = "IntendedVAF", 
                           col_value = "ObsVAF", 
                           accuracy_goal = 50)
# get LoQ values
results$LoQ_values

# get total error table
results$total_error_table |> 
  knitr::kable()

```

We must choose an initial accuracy goal, but we can easily use the plot and the table with total error percentage to select a new `accuracy_goal` if we would like to do so. 


Additional features of this function:

- If there are more than three reagent lots, the function automatically pools them (per CLSI guidelines)
    - Can set `col_lot = NULL` to pool all lots
    - Can set `always_sep_lots = TRUE` to evaluate all lots separately

